Verification of the CVM algorithm with a New Recursive Analysis Technique

### 2025-02-06 paper/main.tex
Paper   - Lots of edits   - Add formalized version of algorithm
Paper: Section on the original proof.
Minor edits   - Add classification, keywords.   - Minor typos.

\usepackage{bigints}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

@@

@@

\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

@@

@@

>>>>

\usepackage{booktabs}
\usepackage{mdframed}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

@@

\definecolor{shadecolor}{gray}{0.95}
\newenvironment{isabelle_cm}{\begin{mdframed}[backgroundcolor=shadecolor,nobreak=true,linewidth=0]\begin{isabelle}}{\end{isabelle}\end{mdframed}}%

@@

\ccsdesc[500]{Theory of computation~Logic and verification}
\ccsdesc[500]{Theory of computation~Higher order logic}
\ccsdesc[500]{Mathematics of computing~Probabilistic algorithms}
\ccsdesc[500]{Theory of computation~Pseudorandomness and derandomization}

\keywords{Formal Verification, Isabelle/HOL, Randomized Algorithms, Distinct Elements}

@@

\input{sec/formalization_neg_dep}

@@

\input{sec/related_work}

\input{sec/conclusion}

### 2025-02-06 paper/sec/formalization.tex
Paper: Edit intro a bit more.
Avoiding reference to the simplicity of the original proof, which contradicts the main point of the paper.
Keep most of the changes previous commit.
Adding summary of sections.
Minor edits   - Add classification, keywords.   - Minor typos.

\section{Formalization of the CVM algorithm}

@@

\medskip
The formalization contains four theory files. The file CVM{\textunderscore}Preliminary definitions and lemmas, such as properties about the \isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} operation we introduced.
Of course, we do not verify \cref{alg:cvm} and \cref{alg:cvm_new} separately,
as most of the proof is identical.
The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, wiht an abstract subsampling operation, which is required to fullfill \cref{eq:subsample_condition}.
The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
It should be noted that only the last file, for the new unbiased algorithm, dependends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
The total number of lines required for the verification of the original algorithm is 1004 lines.
This is in contrast to the original proof, for which the verification required \todo{x} lines.

>>>>

\section{Formalization of the CVM algorithm}\label{sec:formalization}

@@

\medskip
The formalization contains four theory files. The file CVM{\textunderscore}Preliminary definitions and lemmas, such as properties about the \isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} operation we introduced.
Of course, we do not verify \cref{alg:cvm} and \cref{alg:cvm_new} separately,
as most of the proof is identical.
The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{eq:subsample_condition}.
The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
It should be noted that only the last file, for the new unbiased algorithm, dependends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
The total number of lines required for the verification of the original algorithm is 1004 lines.
This is in contrast to the original proof, for which the verification required \todo{x} lines.

### 2025-02-06 paper/sec/intro.tex
Minor edits   - Add classification, keywords.   - Minor typos.

%Curiously, our formalization takes a substantially different route compared to the original proof~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191}.
Briefly, the pen-and-paper analysis for CVM relies on a sequence of transformations of the algorithm.
The reason for these transformations is that standard methods for analyzing randomized algorithms, such as Chernoff--Hoeffding bounds, usually make statements about independent random variables.
Yet, for \cref{alg:cvm} the random operations are far from being independent.\footnote{This was an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022}; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
The original error serves as a side motivation for this work.}
For example, in line 3 the Bernoulli distribution is sampled for a value $p$, which itself depends on previous random operations; similarly, the subsampling step in line 9 is only applied if the buffer is full, which also depends on previous random operations.
The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, from which the desired results about the original algorithm can be deduced.
Indeed, it seems impossible to analyze~\cref{alg:cvm} directly using known methods.

In this paper, we present a new technique for analyzing randomized algorithms.
The technique yields a direct and substantially more general proof of the CVM algorithm.
It is very similar to how deterministic algorithms are verified using a loop invariant.
The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
\[
  \expect [ h ] \leq h(c)
\]
where $h$ is allowed to range over a class of functions, and the expectation is taken over the distribution of the state of the algorithm after consuming each stream element.
By first establishing such an invariant for~\cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a simple but substantially more general proof of the CVM algorithm.
We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to interactive theorem proving.

>>>>

%Curiously, our formalization takes a substantially different route compared to the original proof~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191}.
The pen-and-paper analysis for CVM relies on a sequence of transformations of the algorithm.
The reason for these transformations is that standard methods for analyzing randomized algorithms, such as Chernoff--Hoeffding bounds, usually make statements about independent random variables.
Yet, for \cref{alg:cvm} the state variables are far from being independent.\footnote{There is an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022} that the indicator functions for elements in $\chi$ are; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
The original error serves as a side motivation for this work.}
For example, in line 3 the Bernoulli distribution is sampled for a value $p$, which itself depends on previous random operations; similarly, the subsampling step in line 9 is only applied if the buffer is full, which also depends on previous random operations.
The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, from which the desired results about the original algorithm can be deduced.
Indeed, it seems impossible to analyze \cref{alg:cvm} directly using known methods.

In this paper, we present a new technique for analyzing randomized algorithms.
The technique yields a direct and substantially more general proof of the CVM algorithm.
It is very similar to how deterministic algorithms are verified using a loop invariant.
The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
\[
  \expect [ h ] \leq h(c)
\]
where $h$ is allowed to range over a class of functions, and the expectation is taken over the distribution of the state of the algorithm after consuming each stream element.
By first establishing such an invariant for \cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a simple but substantially more general proof of the CVM algorithm.
We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to interactive theorem proving.
